---
title: "Final Project"
output: html_notebook
authors: "Evelyn Murray and Joseph Pevner"
---

```{r}
library(mosaic)
library(tidyverse)
library(lubridate)
library(DataComputing)
library(rvest)
library(broom)
```

## Research Focus:

As COVID-19 spreads at an alarming rate, a pressing question at a global scale emerges-- what factors of a country contribute to the spread of Coronavirus. Most often cited factors of spread are 



How do different aspects of determining how developed a country is (such as birth rate, population, death rate, etc.) manifest themselves in the spread of COVID-19.  Essentially, how does COVID-19 show different levels of impact on different countries?





























## Data Access

Reading in the Data:


Data Source 1: COVID
```{r}
COVID <- read.csv(file = "total-covid-cases-deaths-per-million.csv")
COVID
```

```{r}
COVID %>%
  nrow()
```
```{r}
COVID %>%
  names()
```
```{r}
COVID %>%
  head()
```



Data Source 2: CountryData
```{r}
CountryData
```

```{r}
CountryData %>%
  nrow()
```
```{r}
CountryData %>%
  names()
```
```{r}
CountryData %>%
  head()
```



Data Source 3: Continents
```{r}
Continents <- read.csv(file = "countries and continents.csv")
Continents
```
```{r}
Continents %>%
  nrow()
```
```{r}
Continents %>%
  names()
```
```{r}
Continents %>%
  head()
```




### Data Wrangling of COVID Dataset

```{r}
COVID
```


```{r}
TidyCOVID <- COVID %>%
  rename(country = total.covid.cases.deaths.per.million ) %>%
  rename( Code = X ) %>%
  rename(Date = X.1 ) %>%
  rename(DeathsPerMillion = X.2) %>%
  rename(CasesPerMillion = X.3) %>%
  filter(row_number() > 1) %>%
  subset(select = c(1,2,3,4,5)) %>%
  mutate( country = as.character(country) ) %>%
  mutate( Code = as.character(Code) ) %>%
  mutate(Date = mdy(Date)) %>%
  mutate(DeathsPerMillion = as.integer(DeathsPerMillion) - 1) %>%
  mutate(CasesPerMillion = as.integer(CasesPerMillion) - 1)

TidyCOVID

```


```{r}
TidyCOVID

```





```{r}

February1 <-
  TidyCOVID %>%
  filter(Date == "2020-02-01") %>%
  subset(select = c(1,4,5)) %>%
  rename(DeathsPerMillionFeb1 = DeathsPerMillion) %>%
  rename(CasesPerMillionFeb1 = CasesPerMillion)

February29 <- 
  TidyCOVID %>%
  filter(Date == "2020-02-29") %>%
  subset(select = c(1,4,5)) %>%
  rename(DeathsPerMillionFeb29 = DeathsPerMillion) %>%
  rename(CasesPerMillionFeb29 = CasesPerMillion)

FebGrow <-
  full_join(February1, February29, by = c("country")) %>%
  mutate("CasesPerMillionFebGrowth" = CasesPerMillionFeb29 - CasesPerMillionFeb1) %>%
  mutate("DeathsPerMillionFebGrowth" = DeathsPerMillionFeb29 - DeathsPerMillionFeb1)




March1 <-
  TidyCOVID %>%
  filter(Date == "2020-03-01") %>%
  subset(select = c(1,4,5)) %>%
  rename(DeathsPerMillionMarch1 = DeathsPerMillion) %>%
  rename(CasesPerMillionMarch1 = CasesPerMillion)

March31 <- 
  TidyCOVID %>%
  filter(Date == "2020-03-31") %>%
  subset(select = c(1,4,5)) %>%
  rename(DeathsPerMillionMarch31 = DeathsPerMillion) %>%
  rename(CasesPerMillionMarch31 = CasesPerMillion)

MarchGrow <-
  full_join(March1, March31, by = c("country")) %>%
  mutate("CasesPerMillionMarchGrowth" = CasesPerMillionMarch31 - CasesPerMillionMarch1) %>%
  mutate("DeathsPerMillionMarchGrowth" = DeathsPerMillionMarch31 - DeathsPerMillionMarch1)



December31 <-
  TidyCOVID %>%
  filter(Date == "2019-12-31") %>%
  subset(select = c(1,4,5)) %>%
  rename(DeathsPerMillionDecember31 = DeathsPerMillion) %>%
  rename(CasesPerMillionDecember31 = CasesPerMillion)


April5 <-
  TidyCOVID %>%
  filter(Date == "2020-04-05") %>%
  subset(select = c(1,4,5)) %>%
  rename(DeathsPerMillionApril5 = DeathsPerMillion) %>%
  rename(CasesPerMillionApril5 = CasesPerMillion)

TotalGrow <-
  full_join(December31, April5, by = c("country")) %>%
  mutate("CasesPerMillionTotalGrowth" = CasesPerMillionApril5 - CasesPerMillionDecember31) %>%
  mutate("DeathsPerMillionTotalGrowth" = DeathsPerMillionApril5 - DeathsPerMillionDecember31)


```

```{r}


FebGrow <-
  FebGrow %>%
  subset(select = c(1, 6, 7))

MarchGrow <-
  MarchGrow %>%
  subset(select = c(1, 6, 7))

TotalGrow <-
  TotalGrow %>%
  subset(select = c(1, 6, 7))


Growth <-
  left_join(FebGrow,MarchGrow) %>%
  left_join(TotalGrow)

Growth

```










### Data Wrangling of Continents
```{r}
Continents

```

```{r}
TidyCont <-
  Continents %>%
  filter(row_number() > 2) %>%
  subset(select = c(1,23)) %>%
  rename(country = name ) %>%
  mutate( country = as.character(country) ) %>%
  mutate( Continent = as.character(Continent) )

TidyCont


```


```{r}
COVIDwCont <-
  left_join(TidyCOVID, TidyCont, by = "country")
```

## Data Visualization
```{r}
DataMix <- left_join(COVIDwCont, CountryData, by = "country")
```

JOEY: I think this is a good one to start out with, but if there is some way that you could make this more clear (color code, etc.) that would be great.  This is a good start to our data.
```{r}

ggplot(data = DataMix,aes(x = Date, y = CasesPerMillion)) +
  geom_point() +
  facet_wrap(~Continent)


```

JOEY: I like this graph because it visualizes one of the factors that contributes to the development classification.  Can we add in a color code or something that would represent one more factor, such as different countries or regions of the world.  The more that we can do with this one, the better.

```{r}

ggplot(data= DataMix, aes(x= Date, y = DeathsPerMillion)) +
  geom_point() +
  facet_wrap(~Continent)


```





```{r}

GrowthMix <-left_join(Growth, CountryData, by = "country")


```



```{r}
ggplot(data=GrowthMix, aes(x = airports,y = CasesPerMillionTotalGrowth)) +
  geom_point() +
  xlim(0,5000)

```

```{r}

ggplot(data=GrowthMix, aes(x = health,y = CasesPerMillionTotalGrowth)) +
  geom_point()

```
```{r}

ggplot(data=GrowthMix, aes(x = health,y = DeathsPerMillionTotalGrowth)) +
  geom_point()

```

```{r}
GrowthMix %>%
  arrange(desc(CasesPerMillionTotalGrowth))
```



```{r}

ShowChange <-
  GrowthMix %>%
  select(c(1, 6, 7, 9)) %>%
  filter(country != "Africa",
         country != "Asia",
         country != "Europe",
         country != "North America",
         country != "Oceania",
         country != "South America",
         country != "World"
         ) %>%
  mutate("TotalGrowth" = (CasesPerMillionTotalGrowth * round(pop/1000000, digits = 0))) %>%
  mutate("TotalDeath" = (DeathsPerMillionTotalGrowth * round(pop/1000000, digits = 0)))

ShowChange %>%
  arrange(desc(TotalDeath))
```





```{r}

WorldMap(ShowChange, key = country, fill = TotalGrowth)

```



