---
title: "Final Project Preliminary EDA"
output: html_notebook
authors: Evelyn Murray and Joseph Pevner
---
##Guiding Question (i.e., Research Question):
How do different aspects of determining how developed a country is (such as birth rate, population, death rate, etc.) manifest themselves in the spread of COVID-19.  Essentially, how does COVID-19 show different levels of impact on different countries?

##Data Source Exploration
1. Dataset 1 - Total COVID Cases and Deaths per Million
This dataset was accessed through the "Our World in Data" website.  Our World in Data collects their information on health from the European CDC, who reports on the progression of the COVID-19 pandemic for the whole world.  They shifted away from using data from the World Health Organization because they found many errors in reporting as well as a more significant lag time in data publishing.  This data is collected to stengthen the world's "defense against infectious diseases" and to keep the most updated tracking possible on the COVID-19 pandemic.  
Each case in this dataset represents an individual day in the collection of data on COVID-19.  In total, there are 9,487 individual cases.   (JOEY, need code to show how many individual cases there are)
```{r}

```
From this dataset, we will most likely be utilizing all of the variables which include: entity (country), code (country code), date (currently not in m/d/y format), total confirmed deaths per million, and total confirmed cases per million. (JOEY, need code to show the column names here)
```{r}

```
2. Dataset 2 - CountryData
This dataset was accessed through the DataComputing Package.  This data is collected and maintained by the CIA factbook.  It was collected so that there is some type of database that contains various "geographic, demographic, and economic data" for each country.  
Each case in this dataset represents an individual country, and there are 256 countries displayed in the dataset in total. (JOEY, need code to show how many individual cases there are)
```{r}

```
From this dataset, we will be utilizing many of the variables that relate to the classifications of socio-economic development.  For example, we will definitely have use for "pop" (the number of people), "growth" (the growth rate per year), "birth" (birth rate per 1000 people), and death (death rate per 1000 people). (JOEY, need code here to show the column names)
```{r}

```


## Intuitive Exploration
Explore intuition related to the research question
Create some informative plots and summary statistics 
Describe preliminary observations and intuition about the research question

