---
title: "Final Project Preliminary EDA"
output: html_notebook
authors: "Evelyn Murray and Joseph Pevner"
---

ONCE YOU GO THROUGH AND MAKE THE GRAPHS, I CAN GO THROUGH AND EXPLAIN EVERYTHING AND MAKE IT PRETTY :)
## Guiding Question (i.e., Research Question):

How do different aspects of determining how developed a country is (such as birth rate, population, death rate, etc.) manifest themselves in the spread of COVID-19.  Essentially, how does COVID-19 show different levels of impact on different countries?

Reading in the Data:
JOEY: This code still won't run on my computer, so I need you to either run the code or change it so that it can be run on other computers
```{r}

COVID <- read.csv("/Users/joeypevner/Final-Project/total-covid-cases-deaths-per-million.csv") %>%
  rename(c("country" = "total.covid.cases.deaths.per.million") ) %>%
  rename(c("Code" = "X") ) %>%
  rename(c("Date" = "X.1") ) %>%
  rename(c("DeathsPerMillion" = "X.2")) %>%
  rename(c("CasesPerMillion" = "X.3")) %>%
  filter(row_number() > 1) %>%
  subset(select = c(1,2,3,4,5)) %>%
  mutate( country = as.character(country) ) %>%
  mutate( Code = as.character(Code) ) %>%
  mutate(Date = mdy(Date)) %>%
  mutate(DeathsPerMillion = as.integer(DeathsPerMillion) - 1) %>%
  mutate(CasesPerMillion = as.integer(CasesPerMillion) - 1)

COVID

```

## Data Sources
JOEY: Access to all data sources is contained within the analysis. 

I think this means that we need a link or a way for people to be able to access the data, but maybe I am misunderstanding this.  Let me know what you think.




## Data Visualization
```{r}
All <- left_join(COVID, CountryData)
```

JOEY: I think this is a good one to start out with, but if there is some way that you could make this more clear (color code, etc.) that would be great.  This is a good start to our data.
```{r}

ggplot(data = All,aes(x = Date, y = CasesPerMillion)) +
  geom_point() 


```

JOEY: I like this graph because it visualizes one of the factors that contributes to the development classification.  Can we add in a color code or something that would represent one more factor, such as different countries or regions of the world.  The more that we can do with this one, the better.

```{r}

ggplot(data=Growth, aes(x=health, y = DeathsPerMillionMarchGrowth)) +
  geom_point() 


```

JOEY:  I also liked this graph, but we need to remove the outlier that scews it and makes the scale hard to see a relationship.  It is the one with an X-value of over 10,000

```{r}
ggplot(data=Growth, aes(x = airports,y = CasesPerMillionMarchGrowth)) +
  geom_point()

```

JOEY: I think one more graph here will be sufficient. Can we do one that compares the relationship between Cases per Million, Deaths per million, and population!!! Maybe something like a cloropleth graph (I think thats what I'm thinking of)?
```{r}

```

