---
title: "Final Project Preliminary EDA"
output: html_notebook
authors: Evelyn Murray and Joseph Pevner
---


## Guiding Question (i.e., Research Question):

How do different aspects of determining how developed a country is (such as birth rate, population, death rate, etc.) manifest themselves in the spread of COVID-19.  Essentially, how does COVID-19 show different levels of impact on different countries?


## Data Source Exploration

1. Dataset 1 - Total COVID Cases and Deaths per Million
This dataset was accessed through the "Our World in Data" website.  Our World in Data collects their information on health from the European CDC, who reports on the progression of the COVID-19 pandemic for the whole world.  They shifted away from using data from the World Health Organization because they found many errors in reporting as well as a more significant lag time in data publishing.  This data is collected to stengthen the world's "defense against infectious diseases" and to keep the most updated tracking possible on the COVID-19 pandemic.  
Each case in this dataset represents an individual day in the collection of data on COVID-19.  In total, there are 9,487 individual cases.  
```{r}

COVID <- read.csv("/Users/joeypevner/Final-Project/total-covid-cases-deaths-per-million.csv") %>%
  rename(c("country" = "total.covid.cases.deaths.per.million") ) %>%
  rename(c("Code" = "X") ) %>%
  rename(c("Date" = "X.1") ) %>%
  rename(c("DeathsPerMillion" = "X.2")) %>%
  rename(c("CasesPerMillion" = "X.3")) %>%
  filter(row_number() > 1) %>%
  subset(select = c(1,2,3,4,5)) %>%
  mutate( country = as.character(country) ) %>%
  mutate( Code = as.character(Code) ) %>%
  mutate(Date = mdy(Date)) %>%
  mutate(DeathsPerMillion = as.integer(DeathsPerMillion) - 1) %>%
  mutate(CasesPerMillion = as.integer(CasesPerMillion) - 1)

COVID

```



```{r}

COVID%>%
  nrow()

```





From this dataset, we will most likely be utilizing all of the variables which include: entity (country), code (country code), date (currently not in m/d/y format), total confirmed deaths per million, and total confirmed cases per million.
```{r}
COVID %>%
  names()
```





2. Dataset 2 - CountryData
This dataset was accessed through the DataComputing Package.  This data is collected and maintained by the CIA factbook.  It was collected so that there is some type of database that contains various "geographic, demographic, and economic data" for each country.  
Each case in this dataset represents an individual country, and there are 256 countries displayed in the dataset in total. 
```{r}
CountryData
```


```{r}


CountryData %>%
  nrow()


```


From this dataset, we will be utilizing many of the variables that relate to the classifications of socio-economic development.  For example, we will definitely have use for "pop" (the number of people), "growth" (the growth rate per year), "birth" (birth rate per 1000 people), and death (death rate per 1000 people). 
```{r}

CountryData %>%
  names()


```


## Intuitive Exploration

```{r}
All <- left_join(COVID, CountryData)
```


```{r}

ggplot(data = All,aes(x = Date, y = CasesPerMillion)) +
  geom_point() 


```
This graph shows that the total cases per million have increased significantly since January of 2020.  While this is rather intuitive data, we will go into more indepth analysis about how different countries have progressed since January, and which are still experiencing a high growth rate in number of cases.
```{r}

February1 <-
  COVID %>%
  filter(Date == "2020-02-01") %>%
  subset(select = c(1,4,5)) %>%
  rename(c("DeathsPerMillionFeb1" = "DeathsPerMillion")) %>%
  rename(c("CasesPerMillionFeb1" = "CasesPerMillion"))

February29 <- 
  COVID %>%
  filter(Date == "2020-02-29") %>%
  subset(select = c(1,4,5)) %>%
  rename(c("DeathsPerMillionFeb29" = "DeathsPerMillion")) %>%
  rename(c("CasesPerMillionFeb29" = "CasesPerMillion"))

FebGrow <-
  full_join(February1, February29, by = c("country")) %>%
  mutate("CasesPerMillionFebGrowth" = CasesPerMillionFeb29 - CasesPerMillionFeb1) %>%
  mutate("DeathsPerMillionFebGrowth" = DeathsPerMillionFeb29 - DeathsPerMillionFeb1)




March1 <-
  COVID %>%
  filter(Date == "2020-03-01") %>%
  subset(select = c(1,4,5)) %>%
  rename(c("DeathsPerMillionMarch1" = "DeathsPerMillion")) %>%
  rename(c("CasesPerMillionMarch1" = "CasesPerMillion"))

March31 <- 
  COVID %>%
  filter(Date == "2020-03-31") %>%
  subset(select = c(1,4,5)) %>%
  rename(c("DeathsPerMillionMarch31" = "DeathsPerMillion")) %>%
  rename(c("CasesPerMillionMarch31" = "CasesPerMillion"))
```
```{r}
MarchGrow <-
  full_join(March1, March31, by = c("country")) %>%
  mutate("CasesPerMillionMarchGrowth" = CasesPerMillionMarch31 - CasesPerMillionMarch1) %>%
  mutate("DeathsPerMillionMarchGrowth" = DeathsPerMillionMarch31 - DeathsPerMillionMarch1)


```

```{r}


FebGrow <-
  FebGrow %>%
  subset(select = c(1, 6, 7))

MarchGrow <-
  MarchGrow %>%
  subset(select = c(1, 6, 7))

Growth <-
  All %>%
  left_join(FebGrow) %>%
  left_join(MarchGrow)

Growth

```
At the end of this table, you can see the addition of our columns called CasesPerMillionFebGrowth, DeathsPerMillionFebGrowth, CasesPerMillionMarchGrowth and DeathsPerMillionMarchGrowth.  These show the change in the variables Deaths Per Million and Cases Per Million to show the change from the first day of the month to the last day of the month.  This data can be further analyzed in many ways that will allow us to see the progression of the virus against different traits of a county.



```{r}


ggplot(data = Growth, aes(x = obesity,y = CasesPerMillionFebGrowth)) + 
  geom_point() 


```
In this graph, you can see that there is little to no direct relationship between the growth in cases in a country and the obesity in that country.  This determines that these two will not be futher considered and explored in our analysis of the two datasets.

```{r}
ggplot(data=Growth, aes(x = airports,y = CasesPerMillionMarchGrowth)) +
  geom_point()

```
This graph would most likely need some manipulation to either remove or alter the effect of the outlier.  However, this graph is meant to show the relationship between the number of airports in a country and their growth in COVID_19 cases throughout the month of March.  This trend most likely would be very positive in the first two weeks of March (when travel was much less restricted) than in the last two weeks.  This would be a trend that would be interesting to further investigate for our analysis.




```{r}

ggplot(data=Growth,aes(x = CasesPerMillionFebGrowth,y = CasesPerMillionMarchGrowth)) +
  geom_point() 


```
This graph shows the negative trend between the CasesPerMillion Growths between February and March.  That means that, for most countries, the higher the case per million growth in February, the lower the case per million growth in March.  The opposite is also true, where the higher the case per million March growth, the lower the case per million February growth.  This is most likely becuase those countries who had higher rates in February took more restrictive measures during this month, while those who had lower rates had not made those same adjustments yet.


```{r}

ggplot(data=Growth, aes(x=health, y = DeathsPerMillionMarchGrowth)) +
  geom_point() 


```
This final graph shows that there is not necessarily a clean relationship between the health spending as a % of the GDP and the deaths per million people growth in the month of March.  Countries with relatively high spending saw both high and low deaths per million, as did those with low.  This graph could possibility be altered with the addition of other factors to make it more meaningful, but right now, this lack of relationship is not particuluarly helpful in our analysis.

