---
title: "Final Project Rough Draft"
output: html_notebook
authors: "Evelyn Murray and Joseph Pevner"
---

```{r}
library(mosaic)
library(tidyverse)
library(lubridate)
library(DataComputing)
library(rvest)
```

ONCE YOU GO THROUGH AND MAKE THE GRAPHS, I CAN GO THROUGH AND EXPLAIN EVERYTHING AND MAKE IT PRETTY :)
## Guiding Question (i.e., Research Question):

How do different aspects of determining how developed a country is (such as birth rate, population, death rate, etc.) manifest themselves in the spread of COVID-19.  Essentially, how does COVID-19 show different levels of impact on different countries?


## Data Access

Reading in the Data:



Data Source 1: COVID
```{r}
COVID <- read.csv(file = "total-covid-cases-deaths-per-million.csv")
COVID
```

```{r}
COVID %>%
  nrow()
```
```{r}
COVID %>%
  names()
```
```{r}
COVID %>%
  head()
```



Data Source 2: CountryData
```{r}
CountryData
```

```{r}
CountryData %>%
  nrow()
```
```{r}
CountryData %>%
  names()
```
```{r}
CountryData %>%
  head()
```






### Data Wrangling of COVID Dataset

```{r}
COVID
```


```{r}
TidyCOVID <-
  COVID %>%
  rename(c("country" = "total.covid.cases.deaths.per.million") ) %>%
  rename(c("Code" = "X") ) %>%
  rename(c("Date" = "X.1") ) %>%
  rename(c("DeathsPerMillion" = "X.2")) %>%
  rename(c("CasesPerMillion" = "X.3")) %>%
  filter(row_number() > 1) %>%
  subset(select = c(1,2,3,4,5)) %>%
  mutate( country = as.character(country) ) %>%
  mutate( Code = as.character(Code) ) %>%
  mutate(Date = mdy(Date)) %>%
  mutate(DeathsPerMillion = as.integer(DeathsPerMillion) - 1) %>%
  mutate(CasesPerMillion = as.integer(CasesPerMillion) - 1)

```


```{r}
TidyCOVID

```




## Data Visualization
```{r}
DataMix <- left_join(TidyCOVID, CountryData, by = "country")
```

JOEY: I think this is a good one to start out with, but if there is some way that you could make this more clear (color code, etc.) that would be great.  This is a good start to our data.
```{r}

ggplot(data = DataMix,aes(x = Date, y = CasesPerMillion)) +
  geom_point() 


```

JOEY: I like this graph because it visualizes one of the factors that contributes to the development classification.  Can we add in a color code or something that would represent one more factor, such as different countries or regions of the world.  The more that we can do with this one, the better.

```{r}

ggplot(data=Growth, aes(x=health, y = DeathsPerMillionMarchGrowth)) +
  geom_point() 


```

JOEY:  I also liked this graph, but we need to remove the outlier that scews it and makes the scale hard to see a relationship.  It is the one with an X-value of over 10,000

```{r}
ggplot(data=Growth, aes(x = airports,y = CasesPerMillionMarchGrowth)) +
  geom_point()

```

JOEY: I think one more graph here will be sufficient. Can we do one that compares the relationship between Cases per Million, Deaths per million, and population!!! Maybe something like a cloropleth graph (I think thats what I'm thinking of)?
```{r}

```


